---
title: "Precision Dosing Summary Report"
author: "Saarland University"
version: "0.1"
linkcolor: "blue"
output:
  pdf_document:
    number_sections: yes
    fig_caption: yes
    keep_tex: no
    latex_engine: xelatex
    toc: no
    toc_depth: 2
documentclass: article
header-includes:
  - \usepackage{graphicx}
  - \usepackage{fancyhdr}
  - \usepackage{geometry}
  - \geometry{a4paper, left=25mm, right=25mm, top=30mm, bottom=30mm}
  - \usepackage{lipsum}
  - \usepackage[table,xcdraw]{xcolor} 
  - \usepackage{hyperref} 
  - \usepackage[most]{tcolorbox}
  - \usepackage{longtable}
  - \usepackage{booktabs}
  - \usepackage{array}
  - \usepackage{float}
  - \definecolor{turquoise}{RGB}{64,224,208}
  - |
    \fancypagestyle{titlepage}{
      \fancyhf{}
      \fancyhead[R]{\includegraphics[width=7cm]{spm_logo.png}}
      \fancyfoot[C]{\includegraphics[width=5cm]{eu_funding_logo.png}}
      \renewcommand{\headrulewidth}{2pt}
      \renewcommand{\headrule}{{\color{turquoise}\hrule width\headwidth height\headrulewidth}}
      \renewcommand{\footrulewidth}{2pt}
      \renewcommand{\footrule}{{\color{turquoise}\hrule width\headwidth height\footrulewidth}}
    }
  - |
    \fancypagestyle{normal}{
      \fancyhf{}
      \fancyfoot[C]{\includegraphics[width=5cm]{eu_funding_logo.png}}
      \renewcommand{\headrulewidth}{0pt}
      \renewcommand{\footrulewidth}{0pt}
    }
params:
  settings : NULL
  order: NULL
  dose_pk: NULL
  user_data: NULL
---

```{r libraries, echo = FALSE, results = 'asis', error = TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(warning = FALSE)
# define libraries
suppressMessages(library(dplyr))
suppressMessages(library(plotly))
suppressMessages(library(ggplot2))
suppressMessages(library(readr))
suppressMessages(library(knitr))
suppressMessages(library(gt))
suppressMessages(library(kableExtra))
suppressMessages(library(ggplot2))
suppressMessages(library(RColorBrewer))
suppressMessages(library(viridis))
suppressMessages(library(paletteer))
suppressMessages(library(glue))
suppressMessages(library(stringr))
```

```{r functions, echo = F, results = 'asis', error = T}
split_dataframe <- function(df, max_labels = 8) {
  unique_labels <- unique(df$label)
  num_labels <- length(unique_labels)
  split_dfs <- list()

  if (num_labels > max_labels) {
    first_group <- unique_labels[1:max_labels]
    split_dfs[["group_1"]] <- df[df$label %in% first_group, ]
    for (i in seq(max_labels + 1, num_labels, by = max_labels)) {
      next_labels <- unique_labels[i:min(i + max_labels - 1, num_labels)]
      split_dfs[[paste("group_", length(split_dfs) + 1, sep = "")]] <- df[df$label %in% next_labels, ]
    }
  } else {
    split_dfs[["group_1"]] <- df
  }
  return(split_dfs)
}
```
```{r, echo = F, results = 'asis', error = T}
# compound and patient id
compound <- params$user_data$model_family
patient <- params$user_data$id

# patient related information
demographics <- data.frame(
  ID = params$user_data$id,
  Age = params$user_data$age,
  Weight = params$user_data$weight,
  Height = params$user_data$height,
  Sex = params$user_data$sex,
  Ethnicity = params$user_data$ethnicity
)

genetics <- params$user_data$genetics$table

dosing <- params$user_data$doses$table

# clinical data
meas_conc <- params$user_data$clinical_conc$table
meas_conc$Date <- format(meas_conc$Date, format = "%Y-%m-%d")
meas_conc$`Clock time` <- format(meas_conc$`Clock time`, format = "%H:%M:%S")

# dose calc
dose_pk <- params$dose_pk$data
dose_pk$date_time <- as.POSIXct(dose_pk$date_time, format = "%Y-%m-%d %H:%M:%S")
```

\thispagestyle{titlepage} 
\begin{center} 
\vspace*{4cm} 
{\Huge \textbf{`r rmarkdown::metadata$title` for `r compound`}} \\
\vspace{1.5cm}
{\Large for Patient: `r patient`}

\vspace{1cm}
{Date: `r format(Sys.Date(), "%d.%m.%Y")`} \\
{Report version: `r rmarkdown::metadata$version`}

\vspace{1cm}
{developed by} \\
{Simeon Rüdesheim, Dominik Selzer, Helena Loer, Laura Fuhr, Fatima Marok, and Thorsten Lehr} \\ 
\textbf{Clinical Pharmacy, Saarland University, Germany} \\

\vfill 
\begin{tcolorbox}[colback=red!5!white,colframe=red!75!black]
\textbf{Disclaimer:} This application is provided for informational and educational purposes only. It is not intended to be a medical device or a substitute for professional medical advice, diagnosis, or treatment. The use of this app should not be considered as a substitute for consulting with a qualified healthcare professional. Users are advised not to make clinical decisions based solely on the results or information provided by this app. Always seek the advice of a healthcare professional for any medical concerns or questions. \\
\end{tcolorbox} 

Grant No. 101057639 

\end{center}

\newpage

\pagestyle{normal}

# General Information
This application is driven by  physiologically-based pharmacokinetic (PBPK) models and predicts patient-specific pharmacokinetics of `r compound` taking individual characteristics, such as age, weight, pharmacogenetics and drug-drug interactions into account.

# Input
## Patient related information 
This section provides an overview of the patient demographics and relevant genetic information for pharmacogenes associated with the drug response provided to the system. Based on respective covariates, the PBPK model predicts the pharmacokinetics of `r compound`.

### Patient Demographics 
```{r demographics-tab, echo = F, results = 'asis', error = T}
kable(
  demographics,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  caption = glue("Demographic information of patient {patient}")
) |>
  kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 12) |>
  row_spec(0, bold = TRUE)
```

### Genetics
```{r genetic-tab, echo = F, results = 'asis', error = T}
if (nrow(genetics) > 0) {
  kable(genetics,
    format = "latex", booktabs = TRUE, escape = FALSE,
    caption = glue("Genetic information of patient {patient}")
  ) |>
    kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 12) |>
    row_spec(0, bold = TRUE)
} else {
  glue("No genetic information provided for patient {patient}.")
}
```

## Dosing Data
The following table displays the administration schedules of `r compound` and the perpetrator drugs provided to the system. This dosing schedule is used to calculate individual `r compound` pharmacokinetics. 

```{r dosing-tab, echo = F, results = 'asis', error = T}
# add unit to dose column name
unit_dose <- params$user_data$doses$value_unit
colnames(dosing)[colnames(dosing) == "Dose"] <- paste0("Dose [", unit_dose, "]")

kable(dosing,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  caption = glue("Dosing information for {compound} and perpetrator drugs")
) |>
  kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 12) |>
  row_spec(0, bold = TRUE)
```

# Results
## Dose simulations
The pharmacokinetics of `r compound` are calculated for different predefined doses. Results are displayed in the following figures and tables.

```{r dosecalc-plot-total, echo = F, results = 'asis', error = T, fig.height=5.5}
cat("<br/>\n\n")

# TODO: REMOVE?
# extract compartment (PK-Sim path)
cmt <- unique(dose_pk$CMT)
parts <- strsplit(cmt, "\\|")[[1]]
# extract compound name from PK-Sim compartment path
drug_name <- parts[3]
sample_info <- parts[4]

# get start of dose adaptation in model
dose_adapt_start_date <- dose_pk$date_time[which(dose_pk$filled == TRUE)[1]]
# format for legend
dose_adapt_start_date_formatted <- format(dose_adapt_start_date, "%Y-%m-%d %H:%M")

# get concentration unit
conc_unit <- params$dose_pk$metaData$PRED$unit

# plot all alternative calculated doses
dose_adapt <- ggplot() +
  geom_line(data = dose_pk, aes(x = date_time, y = DPRED, color = label, linetype = label)) +
  geom_rect(aes(xmin = dose_adapt_start_date, xmax = max(dose_pk$date_time), ymin = -Inf, ymax = Inf),
    fill = "lightgray", alpha = 0.5
  ) +
  labs(
    title = paste(drug_name, "concentration-time profile"),
    x = "Date",
    y = paste0(drug_name, " ", sample_info, " [", conc_unit, "]"),
  ) +
  scale_colour_paletteer_d(palette = "rcartocolor::Safe") +
  guides(color = guide_legend("", ncol = 3), linetype = guide_legend("", ncol = 3)) +
  theme_classic() +
  theme(
    legend.position = "bottom"
  )

# check for measured clinical data and add to plot if available
if (nrow(meas_conc) > 0) {
  dose_adapt <- dose_adapt +
    geom_point(
      data = meas_conc, aes(x = date_time, y = `Concentration [mg/L]`),
      color = "darkred",
      show.legend = FALSE
    )
  print(dose_adapt)
} else {
  print(dose_adapt)
}
```

```{r dosecalc-tab, echo = F, results = 'asis', error = T}
# get pk data
pk_data <- params$dose_pk$summary |>
  select(label, auc, cmax, ctrough, css_average, in_steady_state) |>
  mutate(in_steady_state = ifelse(in_steady_state == TRUE, "yes", "no")) |>
  rename(
    `Steady state` = in_steady_state,
    AUC = auc,
    `Dose` = label,
    `Css average` = css_average,
    `C\\textsubscript{max}` = cmax,
    `C\\textsubscript{trough}` = ctrough,
    `C\\textsubscript{ss, average}` = css_average
  ) |>
  # round for nice visual presentation
  mutate_if(is.numeric, round, 2)

# get units
units <- c(
  "[mg]",
  paste0("[", unique(params$dose_pk$summary$auc_unit), "]"),
  rep(paste0("[", unique(params$dose_pk$summary$c_unit), "]"), 3),
  rep("", 1)
)

# add units to column names
names(pk_data) <- paste(names(pk_data), units)

pk_data$`Dose [mg]` <- gsub("Daily doses: ", "", pk_data$`Dose [mg]`)
pk_data$`Dose [mg]` <- gsub("\\[mg\\]", "", pk_data$`Dose [mg]`)

# generate latex table
kable(pk_data,
  format = "latex", booktabs = TRUE, escape = FALSE,
  col.names = (names(pk_data)),
  caption = glue("Pharmacokinetic parameters of {compound} estimated for different dosing schedules")
) |>
  kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 10) |>
  row_spec(0, bold = TRUE)

cat("<br/>\n\n")
```

```{r dosecalc-fig-subplot, echo = F, results = 'asis', error = T, fig.height=9, fig.width=5.4}
# get model predictions for originally administered dosing schedule
initial_dose <- subset(dose_pk, label == unique(dose_pk$init_label)) |>
  select(Time, date_time, DPRED, init_label)

# rename daily -> original
initial_dose$init_label <- gsub("Daily", "Original", initial_dose$init_label)

# get model predictions for all alternative dosing schedule (- original)
optimized_doses <- subset(dose_pk, label != unique(dose_pk$init_label)) |>
  mutate(dose_num_extr = as.numeric(str_extract(label, "(?<=\\()\\d+"))) |>
  arrange(dose_num_extr)

# split dataframe in packages containing data of max 8 alternative doses (--> max 8 subfigures per plot)
split_result <- split_dataframe(optimized_doses, max_labels = 8)

# generate plots
fig_list <- list()
for (i in seq_len(length(split_result))) {
  # filter data
  dose_set <- split_result[[i]] |>
    # ensure to sort by dose
    arrange(dose_num_extr) |>
    mutate(label = factor(label, levels = unique(label)))
  # generate subplots
  sub_plots <- ggplot(data = dose_set, aes(x = date_time, y = DPRED)) +
    # gray area to indicate start of dose adapation
    annotate("rect",
      xmin = dose_adapt_start_date, xmax = max(dose_set$date_time),
      ymin = -Inf, ymax = Inf, fill = "lightgray", alpha = 0.5
    ) +
    # alternative dose conc-time profile
    geom_line(aes(color = paste0("Alternative doses after ", dose_adapt_start_date_formatted))) +
    # original dose conc-time profile
    geom_line(
      data = initial_dose, aes(x = date_time, y = DPRED, color = init_label),
      linetype = "dashed"
    ) +
    facet_wrap(~label, ncol = 2) +
    labs(
      title = paste(drug_name, "concentration-time profile"),
      x = "Date",
      y = paste0(drug_name, " ", sample_info, " [", conc_unit, "]")
    ) +
    theme_classic() +
    theme(legend.position = "bottom")

  # check for measured clinical data and plot if available
  if (nrow(meas_conc) > 0) {
    # add dosing information to measured concentration
    meas_conc <- meas_conc |>
      mutate(init_label_meas = "Measured Concentration")
    # add measured conc-time points to plot
    sub_plots <- sub_plots +
      geom_point(data = meas_conc, aes(x = date_time, y = `Concentration [mg/L]`, color = init_label_meas)) +
      scale_color_manual(name = "", values = c("#289e47", "darkred", "black")) +
      guides(color = guide_legend(nrow = 2))
  } else {
    sub_plots <- sub_plots +
      scale_color_manual(name = "", values = c("#289e47", "black"))
  }
  fig_list[[i]] <- sub_plots
}

# print all figures
for (p in fig_list) {
  print(p)
  cat("<br/>\n\n")
}
```
